A Privacy Aware Mobile Sensor Application

### 2015-02-05 paper.tex
abstract set added 'organized' section into introduction section 2 modified small error corrections

\begin{abstract}

In our research work we have implemented a privacy aware mobile sensor application [diesen ersten Satz ggf. wieder streichen]. 
In this paper we outline a scenario of a mobile traffic survey and analyse its privacy risks according to a reference model for IT security analysis.
This leads to concrete recommendations for privacy protecting measures that are implemented in the system, for example a Web-based dashboard which allows the users to view, modify and delete all personal data that is stored in the system.

Vorschlag HH: In this article we analyze the privacy aspects of a mobile sensor
application used for recording urban travel patterns as part of a
travel-survey service. This service has been developed and
field-tested within the Live+Gov EU Project. The privacy analysis
follows a structured approach established in
\cite{Grimm:ItSecRefModel}. Eight privacy recommendations are derived,
and have already led to corresponding enhancements of the
travel-survey service.

@@

In this article we study the privacy risks of a travel-survey service
for urban mobility.

@@

Travel surveys are regularly conducted travel agencies which provide
public transportation infrastructure. With these surveys the current
usage of the public transportation system is assessed and the insights
are used for future planning of bus, tram and subway lines, etc..  For
each survey a large number of citizens (e.g. 5000) is asked to keep
track of their travel patterns for a constraint period of time
(e.g. one month). Traditionally the travel patterns have been recorded
manually in the form of travel diaries. Modern smart-phones are
equipped with a wide range of sensors (in particular GPS) that allow
the recording of travel-patterns in a fully automated way.

@@

There are three stakeholders involved in the travel-survey service:
The \emph{citizen}, the \emph{service provider} and the \emph{travel
  agency}. The service provider is the operator of all IT systems and
offers the travel-survey service to the travel agency. Figure
\ref{figure:Live+Gov Vulnerabilities} contains a schematic
visualization of the relation between these stakeholders.

The citizen is a volunteer who is willing to share his personal travel
patterns within the travel-survey event.  He carries a mobile
smart-phone which runs the \emph{sensor collector application}. The
application collects data from various sensors available on modern
smart-phones. In particular accelerometer- and GPS-samples are
collected. Based on the accelerometer data, the application extracts
human activities (e.g. running, walking, standing). Also, the
application can send raw-sensor data back to a central data center.

@@

The \emph{service provider} provides and operates all technical
infrastructure like the data center and the sensor application. He
also operates a web-based \emph{reporting tool} that aggregates
information about travel patterns of citizens within the
travel-survey. The reporting tool is used by the travel agency to gain
insights for planning and accounting of the public transportation
infrastructure.

@@

\subsection{Privacy and Self Data Protection}
More than one hundred years ago, 1890, the American politician Louis Brandis had specified privacy as the right to be left alone \cite{WarrenBrandeis:RightToPrivacy}. From the data processing point of view, this right is best expressed by the absence of information about a person in the mind of others. Indeed, this principle of “data minimization” is still fundamental to modern data protection legislation. The modern principles of data protection are codified by many legal systems in different countries, especially in Europe \cite{EU-directive2}, and by the US save harbor principles. However, modern data protection is more than only the absence of personal data. It is based on the personal right on self-determination over information and communication \cite{BVerfG:census}. This is, for example, very well expressed by Fried’s definition of 1970:
\begin{quote}
	Privacy is not simply an absence of information about us in the minds of others; rather it is the control we have over information about ourselves. \cite{CFried:Privacy}
\end{quote}
In order to realize this requirement in the modern IT world users are provided with user control functions, for example to decide about the access on their data for future use, and to view, modify, and delete unwanted data that had been collected. In its fundamental decision of May 13, 2014, the European Court has convicted Google to accept user demands to delete links to incorrect or irrelevant personal data in the Internet (C-131/12 – Google Spain SL). Such functions would constitute a so-called self-data-protection. It requires users to be highly aware of their rights and how to use them. Complementary, system-data-protection puts the load of data protection enforcement on the data collectors, ideally even without bothering the users. Typical system data protection functions are the deletion of personal data that are not bound to service purpose, the anonymization of personal data for research, and the abstinence from forwarding personal data to third parties. Together, system-data-protection and self-data-protection are supposed to provide a fair balance between the interests of service providers and their users.
In our research work we focus on self-data-protection according to Fried’s user control requirement \cite{CFried:Privacy}. A system data protection point of view would lead to other results, namely to a set of obligations for service providers. This is subject to further research.

@@

In Fried's definition of privacy as control over information, the specification of what constitutes such information remains open. There is a vast amount of information that relates to a person and we need to get a better understanding in order to perform a thorough analysis. To this end we use the categorization by Friedewald, Finn and Wright of 2013 called the Seven Types of Privacy \cite{7ToP}. These are an extension to the four types of privacy by Roger Clarke of 1997 \cite{RClarke:4ToP}. It is important to note, that these categories are not mutually exclusive. For instance a written email is considered personal communication as well as personal data stored on a computer. However, the categories are very helpful for a privacy analysis with a focus on self-data-protection.
The seven types of privacy are as follows: Privacy of the Person (with respect to body functions and body characteristics), Privacy of Behavior and Action, Privacy of Communication, Privacy of Data and Image, Privacy of Thoughts and Feelings, Privacy of Location and Space, Privacy of Association (with persons, e.g. friends, and organizations, e.g. political parties).

@@

\subsubsection{Network Sensors.}
The GSM and WLAN sensors reveal the position of the mobile device and its carrier, when used in connection with external databases.
The both sensors give the exact cell or network, the mobile device has registered with at the current moment.
Frequent connection to one particular network also reveals association, e.g. university networks.

@@

\section{Privacy Analysis}

@@

This section describes the human actors previously introduced within the IT system architecture (\ref{subsubsection:it-systems}). Although the text is only a short description, a list of the most important interests is provided for each actor.

\textbf{Citizen.}
Citizens are persons who use the mobile device as users of the provided software.
Their main motivation for using the software is to gain a higher level of convenience in their daily activities. For example they may have access to real time bus schedules or reports about traffic jams, that help them to avoid long waiting times. Also they generally benefit from improvements of public infrastructure by travel agencies, which is triggered by issue reports.

By using the application, citizens are sharing personal information like name and address, as well as data gathered from mobile sensor with the service provider. This data can be exploited in ways that are harmful to the citizen (cf. \cite{GuardienMassSurveillance}).

Citizens are interested in: physical wellbeing and health, financial profit, convenience, legitimate use of personal data, non-disclosure of personal data to peers of the citizen, and not being monitored.

\textbf{External.}
Externals are persons who do not have privileged access to the IT systems, and are willing to break laws, security constrains and norms in order to promote their interests.

A common interest of an external is financial profit. For example they want to obtain access to critical systems to steal sensitive data or to get the system under their control. Stolen data could simply be sold as is or used for illegitimate purposes, e.g. spam or phishing attacks - or excessive data mining.

In short, externals could be interested in: increase power over citizen, financial profit, political activism, and their social standing.

\textbf{System Provider.}
System Providers operate the technical infrastructure (hardware and software) of the IT System.
They are private companies and legal persons in their own right, but also employ a number of people with diverging interests including administrators, programmer/developer, and a support manager.

As companies, they are interested in gaining financial profit. The financial success of system providers depends on the task complexity of the maintained infrastructure. The complexity of a task has to be in reasonable bounds, so that system providers can complete it within time, with a satisfying quality.

System Providers are interested in: financial profit, manageable complexity, professional excellence, and good working conditions.

\textbf{Local Authority.}
Local authorities are public offices (ministry, agency, department, ...) or other external public entities which act as direct customers of service providers. For example a department for urban mobility, orders a system to better understand usage patterns and make improvement to the urban traffic flow. Such systems are investments, and so naturally travel agencies are interested in a profitable return, like increased ticket sales.

Local authorities are interested in: financial profit, political reputation, business intelligence, and good working conditions.

\subsubsection{Conflicts of Interests}
\label{subsubsection:Conflicts of Interests}
Different actors have different interests which can be in conflict each other. This section outlines the Conflicts of Interests between the actors of the proposed IT system architecture.

The individual interests of all actors is already described in the previous section and are not elaborated any further. The emphasis here is put on prominent existing conflicts, because they provide a foundation for vulnerabilities and subsequent threats.

\textbf{System Complexity vs Privacy.}
System Providers offer a service to Local Authorities, so that Local Authorities can improve their public services. This task in it self has a high technical complexity and operates on privacy sensitive data provided by monitored Citizens.

\textbf{Business Intelligence vs Privacy.}
Local authorities order a monitoring and mining system from system providers as much as possible, which allows them to produce business intelligence for public services. They are interested in the successful usage of their data, although the interest of citizens lies in maintaining control over their data and protecting their rights to privacy.

@@

\textbf{Political Activism vs Reputation of Local Authority.}
Besides monetary reasons, externals can be motivated by political reasons to attack the monitoring and mining system.
Externals can break the system to make a political statement of their own,
or they can steal user data to prove the system insecure.
Both would harm the reputation of travel agencies, who endangered the privacy of the citizens.

@@

\caption{Overview of vulnerabilities of the Live+Gov System; legend see \ref{figure:IT Systems} }
\label{figure:Live+Gov Vulnerabilities}
\end{figure}

@@

\textbf{Inadequate Access Rules.}
The proposed IT system infrastructure has various accesses to privacy sensitive data.
System Provider staff has access to Data Center hardware and software like databases, web-servers and other inspection tools. Local Authority staff has access to the Report Tool. This all enables staff members to have potential access to privacy sensitive information.

@@

The \emph{Potential World Analysis} displays the intended and unintended interactions of the components in the world view.

The intended interactions support the underlying business objectives. Unintended interactions can lead to threats. Obviously this provides a conflict of interest with the victim's business model, to keep the asset away from unauthorized access. From the point of view of the victim, an attack would be an unintended interaction.

The potential world view consists of the following components: Business Objectives, Threats, hances/Risk, and Security Requirements. 
\textbf{Business Objectives} of the system owner were already described in Section \ref{sec:intro} and \ref{subsec:world-analysis}. \textbf{Security Requirement}s are included to our analysis in Section \ref{subsec:priv-rec}.

@@

\textbf{T2. Excessive Data Mining.}
The System Provider and/or the Local Authority secretly extract more private information from the collected data, than the Citizen agreed to. But results of the mining process create no disadvantages for Citizens, because there is no disclosure to third parties. This could be the case for a System Provider, who wants to test a new product and uses the pre-existing data collection.

Thus there are two possible conflicts: \textit{Privacy vs. Financial Profit of Serivce Provider} and  \textit{Privacy vs. Buisness Intelligence of Local Authority}
The threat can be provoked by either lax data handling policies of both System Providers and Local Authorities, or a weak law enforcement of existing supervision.
But the main issue, which can lead to such threats, is again a general \textit{Missing Privacy Awareness}.

\textbf{T3. Data Theft}
An External infiltrates infrastructure in order to steal personal data and sell it on the black market. Also the External might be motivated politically and wants to harm the reputation of the System Provider or the Local Authority. Such a successful attack could harm the reputation of both System Provider and Local Authority. Thus this threat is defined by three conflicts: \textit{Privacy vs. Financial Profit}, \textit{Reputation of Service Provider vs. Political Activism} and \textit{Reputation of Local Authority vs. Political Activism}.
Also this threat describes the classical scenario, where attacks are provoked by \textit{Insecure Infrastructure} (SQL injection) and \textit{Insecure Communication} (Packet Capture).

@@

\textbf{T5. Information Leak}

Like an external person the Data Theft Scenario an employee of the
service provider or the Local Authority has selfish interests to gain
money, make political statements or harm his employer.  In order to
pursue this interest he can steal personal data and sell it or release
it to the public.  The corresponding conflicts of interests are:
\textit{Privacy vs. Financial Profit} of the Employee, \textit{Reputation of Service
  Provider vs. Political Activism} of the Employee and \textit{Reputation of Local
  Authority vs. Political Activism} of the Employee.  The vulnerability constitutes of
the existence of \textit{Unhappy employees} itself and possibly \textit{lax access
rules} that enable the employee to obtain large amounts of data
unnoticed.

\textbf{T6. Social Engineering}
This scenario an external manipulates an employee of a Service
Provider or the Local Authority to leak information to the external
person.  It is thus combination of the Data Theft and Information Leak
scenario.
The conflicts of interest are \textit{Privacy vs. Financial Profit} of
the External, \textit{Reputation of Service Provider vs. Political
  Activism} of the external and \textit{Reputation of Local Authority
  vs. Political Activism} of the external.  The exploited
vulnerabilities are, again, the existence of \textit{Unhappy
  employees} and possibly \textit{lax access rules} that enable the
employee to obtain large amounts of data unnoticed.

@@

For the quantification of the loss in case of a threat scenario we use the following rough calibration between 3 (high) and 0 (none). For the quantification of likeliness the following scale between 4 (always) and 0 (impossible) is used. The quantification of the risk, we add the values for loss and likeliness of the corresponding threats. Note, that loss and
likeliness scales have a logarithmic character, so that addition
of those scales corresponds to multiplication of the usual scales.

The likeliness, loss and the resulting risks assigned to the threats
are discussed in the following paragraphs and summarized in Figure
\ref{fig:risks}.

\textbf{Risk of T1. Insufficient Control Features.}  The occurrence of this
threat is dependent on the design on the system and given in our case,
since we do not give the citizen control over his data once it is
recorded. Therefore the Likeliness is evaluated as $4$ (Always).  The
associated, risk is $1$ Low on our scale, since no direct harm is done to
the citizen by exploiting the data.

Hence the resulting risk is calculated as $4+1 = 5$.

\textbf{Risk of T2. Excessive Data Mining.}
We assess the likeliness of excessive data mining to be $3$ High, since
these kind of analysis can be performed within the walls of the service
provider, without somebody else noticing, and the service provider himself
has an interest in this activity. The associated loss, on the other hand can be substantial (Medium $2$).

Hence the resulting risk is calculated as $3+2 = 5$.

\textbf{Risk of T3. Data Theft.}
The likeliness of a targeted attack by a third party is dependent on
the popularity of the offered service and financial value of the
captured information. Moreover, the amount of manual work required to
infiltrate a custom build system is significantly higher that that of
compromising a standard software solution. In the scenario we assume a
moderate popularity in a single metropolitan area, with around 10.000
users and storage of data of only limited financial value (no
addresses, no payment information). Therefore the likeliness
assessment is $1-2$ (Low-Medium).

The harm of leaked information to a criminal party is $3$ High.
Hence the resulting risk is calculated as $4-5$.

\textbf{Risk of T4. Surveillance.}  In the surveillance scenario an party
related to the citizen, like a company where he is customer of, or a
government agency, seeks to obtain sensitive information from our
service.

The likeliness of such an intrusion is hard to asses, and depends
again on the popularity of the service. If a high popularity is
reached we have recently learned that spying by government agencies is
very likely to occur. The barrier for companies that do not operate
the infrastructure used to transmit the data a surveillance attack is
however very hard to perform. Therefore we assess the likeliness of
the threat with $1-2$ (Low-Medium).

The harm of leaked information to a related party is $3$ High.
Hence the resulting risk is calculated as $4-5$.

\textbf{Risk of T5/6. Information Leak and Social Engineering.} In our scenario we assume that the culture and ethics inside the
service provider company and travel agencies are very high, so that
the information leak scenario has a likeliness of $1$ (Low).

The harm of such an information leaked is $3$ High, so that the
resulting risk is calculated as $4$.

@@

In the preceding section we have identified the main risks for the
users privacy.  In this section we derive recommendations or
requirements for a system that addresses these risks..

In order to address the threat with the highers risk, Insufficient
Control (T1) of the citizen, we need to give the citizen back the
control over its data inside the system. The most direct way to do
this is to provide a web-based \textbf{Privacy Dashboard} (R1) which allows
the citizen to view, edit and delete all information about his person
that is stored inside the system. Also, control over processing and onward transfer of the data to third parties should be given to the user, at least in the form of an opt-out or veto option.

A necessary pre-requirement for effective control of the citizen over
his data is information and comprehension of the intended data
capturing and processing steps. Therefore a \textbf{Privacy Policy} (R2) that
is easily readable and contains all important information is
essential and a legal requirement.

The threat with the second largest risk is (T2) Excessive Data Mining.
Contrary to common belief, it is neither legal nor ethical to process
personal data for by new methods or for new purposes that were not
stated and explained to the citizen at the time of data
collection. Also the common practice of obtaining far-reaching
permissions from the citizens inside the privacy policy is neither an
ethical or legal solution to the problem. To address this threat awareness about the limitations of data
processors inside the company is a key element. As one mean to
establish such a culture of privacy respect, we recommend to prepare
an document called \textbf{Data Handling Guidelines} (R3) intended for
internal use that explains the concrete processing steps and purposes
that are permitted by the citizens. In particular the following information should be provided for each processing task: The name of the controller, purpose of processing,
description of the data categories, recipients of the data if
disclosed, transfer to third countries and a description of security
of processing.

If further processing should be performed, it is necessary to seek
additional permissions from the citizen. A simple email explaining the
planned processing steps, and asking for \textbf{permission} (R4) would be
enough for this purpose. The permission can be given via an embedded
link that shall be followed in order to signal agreement.

An alternative measure to address the risk of excessive data mining is
the \textbf{anonymization} (R5) of data. When all direct- or indirect
links to the identity of the person are removed, no violation of the
citizens privacy caused by arbitrary processing. However removing all
such links is a challenging tasks, and full anonymity is often not
achieved, cf. \cite{krumm2009}.

The protection from threat scenario (T3) Data Theft is a case of
classical \textbf{IT infrastructure security} (R6). The storage and processing infrastructure has
to be secured using firewalls, up-to data software versions and proper
authentication mechanisms.

The protection from threat scenario (T4) Surveillance focuses on the
communication channels. They are target of wiretapping attacks by
intermediaries or externals with access to the communication
infrastructure. Strong \textbf{encryption} (R7) should be used to make
it harder for externals to read the content of the transmitted data.

Threat scenarios (T5) Information Leak and (T6) Social Engineering
target the vulnerability of unhappy employees. Therefore a trustful,
\textbf{healthy company culture} (R8) should be maintained.

The mapping of these recommendations to the threats is summarized in Figure \ref{fig:risks}.
%\end{LGContent}

>>>>

\begin{abstract}
In this article we analyze the privacy aspects of a mobile sensor
application used for recording urban travel patterns as part of a
travel-survey service. This service has been developed and
field-tested within the Live+Gov EU Project. The privacy analysis
follows a structured approach established in
\cite{Grimm:ItSecRefModel}. Eight privacy recommendations are derived,
and have already led to corresponding enhancements of the
travel-survey service.

@@

In this article we study the privacy risks of a travel-survey service
for urban mobility that is further specified in section \ref{sec:ScenarioDescription}. In section \ref{sec:privacydef} below, we will explain our view on privacy. In order to find the right user control functions we will perform a systematic privacy analysis of our sensor data storage and mining infrastructure following the reference model of IT security analysis \cite{Grimm:ItSecRefModel} in section \ref{sec:privacyanalysis} below.

@@

Travel surveys are regularly conducted by transport authorities which provide
public transportation infrastructure. With these surveys the current
usage of the public transportation system is assessed and the insights
are used for future planning of bus, tram and subway lines, etc..  For
each survey a large number of citizens (e.g. 5000) is asked to keep
track of their travel patterns for a constraint period of time
(e.g. one month). Traditionally the travel patterns have been recorded
manually in the form of travel diaries. Modern smart-phones are
equipped with a wide range of sensors (in particular GPS) that allow
the recording of travel-patterns in a fully automated way.

The travel services need the personal mobile data for two reasons: firstly, they want to serve their customers individually, for example by guiding them to proximity stations, to recommend adequate travel connections, to inform about delays, or to sell electronic tickets. For this purpose, the personal data for evaluation are related to the individual customers. Secondly, the travel services want to enhance their system, in that they evaluate the demand of stations, connections, changes, ways and time spent by their customers. For this purpose, personal data are anonymized before evaluation.

@@

There are three stakeholders involved in the travel-survey service:
The \emph{citizen}, the \emph{service provider} and the \emph{transport
  authority}. The service provider is the operator of all IT systems and
offers the travel-survey service to the transport authority. Figure
\ref{figure:Live+Gov Vulnerabilities} contains a schematic
visualization of the relation between these stakeholders.

The \emph{citizen} is a volunteer who is willing to share his personal travel
patterns within the travel-survey event.  He carries a mobile
smart-phone which runs the \emph{sensor collector application}. The
application collects data from various sensors available on modern
smart-phones. In particular accelerometer- and GPS-samples are
collected. Based on the accelerometer data, the application extracts
human activities (e.g. running, walking, standing). Also, the
application can send raw-sensor data back to a central data center.

@@

The \emph{service provider} provides and operates all technical
infrastructure like the data center and the sensor application. He
also operates a web-based \emph{reporting tool} that aggregates
information about travel patterns of citizens within the
travel-survey. The reporting tool is used by the transport authority to gain
insights for planning and accounting of the public transportation
infrastructure.

@@

\subsection{Privacy and Self Data Protection}
1890, the American politician Louis Brandis had specified privacy as the right to be left alone \cite{WarrenBrandeis:RightToPrivacy}. From the data processing point of view, this right is best expressed by the absence of information about a person in the mind of others. Indeed, this principle of “data minimization” is still fundamental to modern data protection legislation. The modern principles of data protection are codified by many legal systems in different countries, especially in Europe \cite{EU-directive2}, and by the US save harbor principles. However, modern data protection is more than only the absence of personal data. It is based on the personal right on self-determination over information and communication \cite{BVerfG:census}. This is, for example, very well expressed by Fried’s definition of 1970:
\begin{quote}
	Privacy is not simply an absence of information about us in the minds of others; rather it is the control we have over information about ourselves. \cite{CFried:Privacy}
\end{quote}
In order to realize this requirement in the modern IT world users are provided with user control functions, for example to decide about the access on their data for future use, and to view, modify, and delete unwanted data that had been collected. In its fundamental decision of May 13, 2014, the European Court has convicted Google to accept user demands to delete links to incorrect or irrelevant personal data in the Internet (C-131/12 – Google Spain SL). Such functions would constitute a so-called self-data-protection. It requires users to be highly aware of their rights and how to use them. Complementary, system-data-protection puts the load of data protection enforcement on the data collectors, ideally even without bothering the users. Typical system data protection functions are the deletion of personal data that are not bound to service purpose, the anonymization of personal data for research, and the abstinence from forwarding personal data to third parties. Together, system-data-protection and self-data-protection are supposed to provide a fair balance between the interests of service providers and their users.
In our research work we focus on self-data-protection according to Fried’s user control requirement \cite{CFried:Privacy}. A system data protection point of view would lead to other results, namely to a set of obligations for service providers. This is subject to further research.

@@

In Fried's definition of privacy as control over information, the specification of what constitutes such information remains open. There is a vast amount of information that relates to a person and we need to get a better understanding in order to perform a thorough analysis. To this end we use the categorization by Friedewald, Finn and Wright of 2013 called the Seven Types of Privacy \cite{7ToP}. These are an extension to the four types of privacy by Roger Clarke of 1997 \cite{RClarke:4ToP}. It is important to note, that these categories are not mutually exclusive. For instance a written email is considered personal communication as well as personal data stored on a computer. However, the categories are very helpful for a privacy analysis with a focus on self-data-protection.
The seven types of privacy are as follows: Privacy of the Person (with respect to body functions), Privacy of Behavior and Action, Privacy of Communication, Privacy of Data and Image, Privacy of Thoughts and Feelings, Privacy of Location and Space, Privacy of Association (with persons, e.g. friends, and organizations, e.g. political parties).

@@

\subsubsection{Network Sensors.}
The GSM and WLAN sensors reveal the position of the mobile device and its carrier, when used in connection with external databases.
The both sensors give the exact cell or network, the mobile device has registered with at the current moment.
Frequent connection to one particular network also reveals association, e.g. university networks. The Bluetooth sensors record lists of the bluetooth clients in the direct neighbourhood.
Since those clients are usually moving, inference of the position is usually not possible.
Instead, bluetooth clients carried by a third person may infringe the Privacy of Association.

@@

\section{Privacy Analysis}
\label{sec:privacyanalysis}

@@

This section describes the human actors previously introduced in section \ref{sec:ScenarioDescription}. Although the text is only a short description, a list of the most important interests is provided for each actor.

\textbf{Citizen.}
Citizens are persons who use the mobile device as users of the provided software.
Their main motivation for using the software is to gain a higher level of convenience in their daily activities. For example they may have access to real time bus schedules or reports about traffic jams, that help them to avoid long waiting times. Also they generally benefit from improvements of public infrastructure by travel agencies, which is triggered by issue reports. By using the application, citizens are sharing personal information like name and address, as well as data gathered from mobile sensor with the service provider. This data can be exploited in ways that are harmful to the citizen (cf. \cite{GuardienMassSurveillance}). Citizens are interested in: physical wellbeing and health, financial profit, convenience, legitimate use of personal data, non-disclosure of personal data to peers of the citizen, and not being monitored.

\textbf{External.}
Externals are persons who do not have privileged access to the IT systems, and are willing to break laws, security constrains and norms in order to promote their interests. A common interest of an external is financial profit. For example they want to obtain access to critical systems to steal sensitive data or to get the system under their control. Stolen data could simply be sold as is or used for illegitimate purposes, e.g. spam or phishing attacks - or excessive data mining. In short, externals could be interested in: increase power over citizen, financial profit, political activism, and their social standing.

\textbf{System Provider.}
System Providers operate the technical infrastructure (hardware and software) of the IT System. They are private companies and legal persons in their own right, but also employ a number of people with diverging interests including administrators, programmer/developer, and a support manager. As companies, they are interested in gaining financial profit. The financial success of system providers depends on the task complexity of the maintained infrastructure. The complexity of a task has to be in reasonable bounds, so that system providers can complete it within time, with a satisfying quality. System Providers are interested in: financial profit, manageable complexity, professional excellence, and good working conditions.

\textbf{Transport Authority.}
 Transport authorities are public offices (ministry, agency, department, ...) or other external public entities which act as direct customers of service providers. For example a department for urban mobility, orders a system to better understand usage patterns and make improvement to the urban traffic flow. Such systems are investments, and so naturally travel agencies are interested in a profitable return, like increased ticket sales. Transport authorities are interested in: financial profit, political reputation, business intelligence, and good working conditions.

\subsubsection{Conflicts of Interests}
\label{subsubsection:Conflicts of Interests}
Different actors have different interests which can be in conflict each other. This section outlines the Conflicts of Interests between the actors of the proposed IT system architecture. The individual interests of all actors is already described in the previous section and are not elaborated any further. The emphasis here is put on prominent existing conflicts, because they provide a foundation for vulnerabilities and subsequent threats.

\textbf{System Complexity vs Privacy.}
System Providers offer a service to Transport Authorities, so that they are able to improve their public services. This task in it self has a high technical complexity and operates on privacy sensitive data provided by monitored Citizens.

\textbf{Business Intelligence vs Privacy.}
Transport authorities order a monitoring and mining system from system providers as much as possible, which allows them to produce business intelligence for public services. They are interested in the successful usage of their data, although the interest of citizens lies in maintaining control over their data and protecting their rights to privacy.

@@

\textbf{Political Activism vs Reputation of Transport Authority.}
Besides monetary reasons, externals can be motivated by political reasons to attack the monitoring and mining system.
Externals can break the system to make a political statement of their own,
or they can steal user data to prove the system insecure.
Both would harm the reputation of travel agencies, who endangered the privacy of the citizens.

@@

\caption{Overview of vulnerabilities of the Live+Gov System }
\label{figure:Live+Gov Vulnerabilities}
\end{figure}

@@

\textbf{Inadequate Access Rules.}
The proposed IT system infrastructure has various accesses to privacy sensitive data.
System Provider staff has access to Data Center hardware and software like databases, web-servers and other inspection tools. Transport authority staff has access to the Report Tool. This all enables staff members to have potential access to privacy sensitive information.

@@

The \emph{Potential World Analysis} displays the intended and unintended interactions of the components in the world view. The intended interactions support the underlying business objectives. Unintended interactions can lead to threats. Obviously this provides a conflict of interest with the victim's business model, to keep the asset away from unauthorized access. From the point of view of the victim, an attack would be an unintended interaction. The potential world view consists of the following components: Business Objectives, Threats, hances/Risk, and Security Requirements. 
\textbf{Business Objectives} of the system owner were already described in Section \ref{sec:intro} and \ref{subsec:world-analysis}. \textbf{Security Requirements} are included to our analysis in Section \ref{subsec:priv-rec}.

@@

\textbf{T2. Excessive Data Mining.}
The System Provider and/or the transport authority secretly extract more private information from the collected data, than the Citizen agreed to. But results of the mining process create no disadvantages for Citizens, because there is no disclosure to third parties. This could be the case for a System Provider, who wants to test a new product and uses the pre-existing data collection. Thus there are two possible conflicts: \textit{Privacy vs. Financial Profit of Serivce Provider} and  \textit{Privacy vs. Buisness Intelligence of Transport Authority} The threat can be provoked by either lax data handling policies of both System Providers and Transport Authorities, or a weak law enforcement of existing supervision. But the main issue, which can lead to such threats, is again a general \textit{Missing Privacy Awareness}.

\textbf{T3. Data Theft}
An External infiltrates infrastructure in order to steal personal data and sell it on the black market. Also the External might be motivated politically and wants to harm the reputation of the System Provider or the Transport Authority. Such a successful attack could harm the reputation of both System Provider and Transport Authority. Thus this threat is defined by three conflicts: \textit{Privacy vs. Financial Profit}, \textit{Reputation of Service Provider vs. Political Activism} and \textit{Reputation of Transport Authority vs. Political Activism}. Also this threat describes the classical scenario, where attacks are provoked by \textit{Insecure Infrastructure} (SQL injection) and \textit{Insecure Communication} (Packet Capture).

@@

\textbf{T5. Information Leak} 
Like an external person the Data Theft Scenario an employee of the
service provider or the transport authority has selfish interests to gain
money, make political statements or harm his employer.  In order to
pursue this interest he can steal personal data and sell it or release
it to the public.  The corresponding conflicts of interests are:
\textit{Privacy vs. Financial Profit} of the Employee, \textit{Reputation of Service
  Provider vs. Political Activism} of the Employee and \textit{Reputation of Transport
  Authority vs. Political Activism} of the Employee.  The vulnerability constitutes of
the existence of \textit{Unhappy employees} itself and possibly \textit{lax access
rules} that enable the employee to obtain large amounts of data
unnoticed.

\textbf{T6. Social Engineering}
This scenario an external manipulates an employee of a Service
Provider or the Transport Authority to leak information to the external
person.  It is thus combination of the Data Theft and Information Leak
scenario.
The conflicts of interest are \textit{Privacy vs. Financial Profit} of
the External, \textit{Reputation of Service Provider vs. Political
  Activism} of the external and \textit{Reputation of Transport Authority
  vs. Political Activism} of the external.  The exploited
vulnerabilities are, again, the existence of \textit{Unhappy
  employees} and possibly \textit{lax access rules} that enable the
employee to obtain large amounts of data unnoticed.

@@

For the quantification of the loss in case of a threat scenario we use the following rough calibration between 3 (high) and 0 (none). For the quantification of likeliness the following scale between 4 (always) and 0 (impossible) is used. The quantification of the risk, we add the values for loss and likeliness of the corresponding threats. Note, that loss and
likeliness scales have a logarithmic character, so that addition
of those scales corresponds to multiplication of the usual scales. The likeliness, loss and the resulting risks assigned to the threats
are discussed in the following paragraphs and summarized in Figure
\ref{fig:risks}.

\textbf{Risk of T1. Insufficient Control Features.}  The occurrence of this
threat is dependent on the design on the system and given in our case,
since we do not give the citizen control over his data once it is
recorded. Therefore the Likeliness is evaluated as $4$ (Always).  The
associated, risk is $1$ Low on our scale, since no direct harm is done to
the citizen by exploiting the data. Hence the resulting risk is calculated as $4+1 = 5$.

\textbf{Risk of T2. Excessive Data Mining.}
We assess the likeliness of excessive data mining to be $3$ High, since
these kind of analysis can be performed within the walls of the service
provider, without somebody else noticing, and the service provider himself
has an interest in this activity. The associated loss, on the other hand can be substantial (Medium $2$). Hence the resulting risk is calculated as $3+2 = 5$.

\textbf{Risk of T3. Data Theft.}
The likeliness of a targeted attack by a third party is dependent on
the popularity of the offered service and financial value of the
captured information. Moreover, the amount of manual work required to
infiltrate a custom build system is significantly higher that that of
compromising a standard software solution. In the scenario we assume a
moderate popularity in a single metropolitan area, with around 10.000
users and storage of data of only limited financial value (no
addresses, no payment information). Therefore the likeliness
assessment is $1-2$ (Low-Medium). The harm of leaked information to a criminal party is $3$ High. Hence the resulting risk is calculated as $4-5$.

\textbf{Risk of T4. Surveillance.}  In the surveillance scenario an party
related to the citizen, like a company where he is customer of, or a
government agency, seeks to obtain sensitive information from our
service. The likeliness of such an intrusion is hard to asses, and depends
again on the popularity of the service. If a high popularity is
reached we have recently learned that spying by government agencies is
very likely to occur. The barrier for companies that do not operate
the infrastructure used to transmit the data a surveillance attack is
however very hard to perform. Therefore we assess the likeliness of
the threat with $1-2$ (Low-Medium). The harm of leaked information to a related party is $3$ High. Hence the resulting risk is calculated as $4-5$.

\textbf{Risk of T5/6. Information Leak and Social Engineering.} In our scenario we assume that the culture and ethics inside the
service provider company and travel agencies are very high, so that
the information leak scenario has a likeliness of $1$ (Low). The harm of such an information leaked is $3$ High, so that the
resulting risk is calculated as $4$.

@@

In this section we derive recommendations or
requirements for a system that addresses these risks. The mapping of these recommendations to the threats is summarized in Figure \ref{fig:risks}.

\textbf{Privacy Dashboard (R1)}. In order to address the threat with the highers risk, Insufficient Control (T1) of the citizen, we need to give the citizen back the
control over its data inside the system. The most direct way to do
this is to provide a web-based overview which allows the citizen to view, edit and delete all information about his person
that is stored inside the system.

\textbf{Privacy Policy (R2)}. A necessary pre-requirement for effective control of the citizen over his data is information and comprehension of the intended data capturing and processing steps. Therefore a policy that
is easily readable and contains all important information is essential and a legal requirement.

\textbf{Data Handling Guidelines (R3)} The threat with the second largest risk is (T2) Excessive Data Mining.
Contrary to common belief, it is neither legal nor ethical to process
personal data for by new methods or for new purposes that were not
stated and explained to the citizen at the time of data
collection. Also the common practice of obtaining far-reaching
permissions from the citizens inside the privacy policy is neither an
ethical or legal solution to the problem. To address this threat awareness about the limitations of data
processors inside the company is a key element. As one mean to
establish such a culture of privacy respect, we recommend to prepare
an guideline intended for internal use that explains the concrete processing steps and purposes that are permitted by the citizens.

\textbf{User Permissions (R4)}. If further processing should be performed, it is necessary to seek additional permissions from the citizen. A simple email explaining the planned processing steps, and asking for permissions would be enough for this purpose. The permission can be given via an embedded link that shall be followed in order to signal agreement.

\textbf{Anonymization (R5)} An alternative measure to address the risk of excessive data mining is the anonymization of data. When all direct- or indirect
links to the identity of the person are removed, no violation of the
citizens privacy caused by arbitrary processing. However removing all
such links is a challenging tasks, and full anonymity is often not
achieved, cf. \cite{krumm2009}.

\textbf{IT Infrastructure Security (R6)}. The protection from threat scenario (T3) Data Theft is a case of
classical IT infrastructure security. The storage and processing infrastructure has
to be secured using firewalls, up-to data software versions and proper
authentication mechanisms.

\textbf{Encryption (R7)} The protection from threat scenario (T4) Surveillance focuses on the
communication channels. They are target of wiretapping attacks by
intermediaries or externals with access to the communication
infrastructure. Strong encryption should be used to make
it harder for externals to read the content of the transmitted data.

\textbf{Relationship (R8)}. Threat scenarios (T5) Information Leak and (T6) Social Engineering
target the vulnerability of unhappy employees. Therefore a trustful,
healthy company culture should be maintained.

%\end{LGContent}
